2.3 Prob Model for PCR Positivity Estimation

### 2024-02-08 2_scripts/2.3 Prob Model for PCR pos.qmd
Updated codes for figures

## Load packages
```{r}
library(dplyr)         # Data management
library(ggplot2)       # ggplot
library(zoo)           # year-month format
library(lmtest)        # for LR test, comparing models
#library(reshape2)     # for melt() function
library(readxl)        # for reading .xlsx data
library(nnet)          # for multinomial log regression
```

## Load dataset
```{r}
# The CBS Ab subset data
load(file = "../1_data/private/CBS_Ab.RData")
load(file = "../1_data/private/RFD4682e.RData")
```

@@

#### bind all those four data by rows
```{r}
dta_all<-dplyr::bind_rows(dta_01s, dta_03s, dta_04, dta_06s)
```

@@

# Model 02 takes sample weights into account
mnlm02<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, 
                 data = dta_all, weights = sample_wt)
summary(mnlm02)

(summary(mnlm01))$coeffcients/summary(mnlm01)$standard.errors
```

@@

@@

>>>>

## Load packages
```{r}
library(dplyr)         # Data management
library(ggplot2)       # ggplot
library(ggpubr)     #ggplot2 QQ plot for normality check
library(zoo)           # year-month format
library(lmtest)        # for LR test, comparing models
#library(reshape2)     # for melt() function
library(readxl)        # for reading .xlsx data
library(nnet)          # for multinomial log regression
```

## Load dataset
```{r}
# The CBS full data
load(file = "../1_data/private/CBS.RData")
table(CBS$province, useNA = 'ifany')

load(file = "../1_data/private/RFD4682e.RData")
```

@@

#### bind all those four data by rows
```{r}
dta_all<-dplyr::bind_rows(dta_01s, dta_03s, dta_04, dta_06s)

# add log quant assay
dta_all$log_roche_n<-log(dta_all$roche_n)    # COI for log_n = 0
dta_all$log_abbott_n<-log(dta_all$abbott_n)  # COI for log)n = 0.3365 (COI = 1.4)

# save the rbinded harmonized data
save(dta_all, file="../1_data/private/Prob Model/dta_all.RData")
```

@@

# Model 02 takes sample weights into account
mnlm02<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, 
                 data = dta_all, weights = sample_wt)
summary(mnlm02)
```
Notes on sample weights: About the target population estimates we have in the study, it is "the proportion of being infected in past 3 months" at a certain time point (each month). So, we can regard it as a point prevalence of the target population. In this case, samples sampled at different months should be unweighted as they contributed independently to each month's point prevalence estimates.

As for the models we have been working on recently, it's a little bit different. The data we used are longitudinal but we used it in a cross-sectional manner: we pooled the data from all time point together then used quant assay data predicting the probability of being infected in the past three months. In this case, we should apply weights. Otherwise, individuals who were sampled multiple times would disproportionately influence the model parameter estimates, so they need to be downweighted.”

```{r}
# model 03: using log-transformed assay

# check the distribution of roche_n (<COI)
ggdensity(dta_all$roche_n[dta_all$roche_n<1])     # mostly normal, but skewed
ggqqplot(dta_all$roche_n[dta_all$roche_n<1])      # same above
shapiro.test(dta_all$roche_n[dta_all$roche_n<1])  # not normal

# check the distribution of log_e roche_n (<COI)
ggdensity(dta_all$log_roche_n[dta_all$log_roche_n<0])     # mostly normal, but skewed
ggqqplot(dta_all$log_roche_n[dta_all$log_roche_n<0])      # same above
shapiro.test(dta_all$log_roche_n[dta_all$log_roche_n<0])  # not normal

# what if log-10 transformed
shapiro.test(log10(dta_all$roche_n[dta_all$roche_n<1]))

# check the positive cases roche-n
ggdensity(dta_all$roche_n[dta_all$roche_n>1])     # mostly normal, but skewed
ggqqplot(dta_all$roche_n[dta_all$roche_n>1])      # same above
shapiro.test(dta_all$roche_n[dta_all$roche_n>1])  # not normal

# log-e transformed, >COI
ggdensity(dta_all$log_roche_n[dta_all$log_roche_n>0])     # mostly normal, but skewed
ggqqplot(dta_all$log_roche_n[dta_all$log_roche_n>0])      # same above
shapiro.test(dta_all$log_roche_n[dta_all$log_roche_n>0])  # not normal
```
```{r}
# Model 02 takes sample weights into account
mnlm03<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n, 
                 data = dta_all, weights = sample_wt)
summary(mnlm03)
```

@@

need to rewrite this section

@@

##============ Part III. Apply Model to CBS Data ================

```{r}
# Add log roche_n to the CBS data
CBS$log_roche_n<-log(CBS$roche_n)

CBS_prob_r<-as.data.frame(predict(mnlm03, newdata = CBS, "probs")) # num of >3 months = 0
mean(CBS_prob_r[,1], na.rm = T)  # 59.97%
mean(CBS_prob_r[,2], na.rm = T)  # 27.70%
mean(CBS_prob_r[,3], na.rm = T)  # 12.33%

CBSs<-cbind.data.frame(select(CBS, pid, sampledate, sex, age, fsa,
                        province, roche_n, abbott_n, roche_s, log_roche_n),
                 CBS_prob_r)
```

#### Make by Prov Plots
```{r}
# Recategorize by regions
table(CBSs$province)

# 12 provinces, recatgorized into five regions
CBSs$region<-rep(NA, length(CBSs$pid))
CBSs$region[CBSs$province %in% c("NL", "PE", "NS", "NB")]<-"Atlantic"
CBSs$region[CBSs$province %in% c("QC", "ON")]<-"Central"
CBSs$region[CBSs$province %in% c("AB", "MB", "SK")]<-"Prairie"
CBSs$region[CBSs$province %in% c("BC")]<-"BC"
CBSs$region[CBSs$province %in% c("NU/NT", "YT")]<-"North"

table(CBSs$province, CBSs$region, useNA = 'ifany')

# Creat sampling months
CBSs$samplemonth<-as.yearmon(CBSs$sampledate)
table(CBSs$samplemonth, useNA = 'ifany')

# Creat a ever inf indicator
CBSs$ever_inf <- 1-CBSs$never_inf
```

#### Creating melt data
```{r}
# create by-month est for each region
tb_bc<-CBSs[CBSs$region=="BC",] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T))
tb_bc$region<-rep("BC", length(tb_bc$n))

tb_at<-CBSs[CBSs$region=="Atlantic",] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T))
tb_at$region<-rep("Atlantic", length(tb_at$n))

tb_ce<-CBSs[CBSs$region=="Central",] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T))
tb_ce$region<-rep("Central", length(tb_ce$n))

tb_pr<-CBSs[CBSs$region=="Prairie",] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T))
tb_pr$region<-rep("Prairie", length(tb_pr$n))

tb_no<-CBSs[CBSs$region=="North",] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T))
tb_no$region<-rep("North", length(tb_no$n))

tb_all<-rbind.data.frame(tb_bc, tb_pr, tb_ce, tb_at, tb_no)
```

#### Make plots by region
```{r}
# Reshape meltted data
tb_melt<-reshape2::melt(select(tb_all, 1,3:5), 
                        id.vars = c("samplemonth", "region"),
                        variable.name = "inf",
                        value.name = "prop")

p01<-tb_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
  geom_point(aes(colour=inf)) +
#  geom_line(aes(colour=inf))  +
  geom_smooth(method = "loess", aes(colour=inf)) +
  theme_bw()

p01 + facet_wrap(~region)
```
## Part IV Apply model to APL data
```{r}
summary(RFD4682_e$N)
# The codebook does not specify what types of anti-N assay is applied, given the number we assume they are using Abbott anti-N assay (small numbers)
```

## Part V Apply model to CANPATH data
```{r}
Canpath <- read.csv("~/COVID-assay-adjust-methods/1_data/private/CANPATH/DAO-543759_ResearcherDataset_Serology_Results_74503par.csv")
Canpath_all <- read.csv("~/COVID-assay-adjust-methods/1_data/private/CANPATH/DAO-543759_ResearcherDataset_Qx_96014par_1125var.csv")

table(Canpath$C1_CITF_ASSAY_ID)

```
